\documentclass[a4paper,twocolumn]{report}

% ---- pdflatex 基本 ----
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{microtype}

% ---- 箱（ascmac の代替）----
\usepackage{tcolorbox}
\tcbuselibrary{breakable}
\tcbuselibrary{skins}
% ascmac の itembox 代替：第1引数＝tcolorboxのキー（任意），第2引数＝タイトル（必須）
% 例）\begin{itembox}[halign title=left]{Problem} ... \end{itembox}
\newtcolorbox{itembox}[1]{%
  enhanced,breakable,
  colback=white,colframe=black,boxrule=0.4pt,arc=2pt,
  title={#1},
}


% ---- 以降はいつも通り ----
\usepackage{minted}
\usepackage{amsmath,amssymb,amsfonts,bm}
\usepackage{siunitx}
\usepackage{graphicx}
\usepackage[version=4]{mhchem}
\usepackage[colorlinks=true,bookmarks=true,bookmarksnumbered=true,
  linkcolor=cyan,urlcolor=blue,citecolor=blue]{hyperref}


\usepackage{color}
\definecolor{red}{rgb}{0.8,0.0,0.0}
\definecolor{gray}{rgb}{0.3,0.3,0.3}
%\usepackage{algorithm}
\usepackage{algorithmic}
%\usepackage{xltxtra}

\newcommand{\thetav}{\boldsymbol{\theta}}
\newcommand{\dv}{\boldsymbol{d}}
\newcommand{\rv}{\boldsymbol{r}}
\newcommand{\ev}{\boldsymbol{e}}
\newcommand{\Ysim}{\boldsymbol{Y}}
\newcommand{\Ysims}{{Y}}
\newcommand{\Yobs}{\boldsymbol{Y}_\mathrm{obs}}
\newcommand{\Yobss}{{Y}_\mathrm{obs}}
\newcommand{\Ysmsim}{\hat{Y}}
\newcommand{\Ysmobs}{\hat{Y}_\mathrm{obs}}


\newcommand{\Nsample}{N_\mathrm{s}}
\newcommand{\isample}{i_\mathrm{s}}
\newcommand{\Nthread}{N_\mathrm{th}}
\newcommand{\ithread}{i_\mathrm{th}}
\newcommand{\Nparticle}{N_\mathrm{p}}
\newcommand{\sobs}{s(\Yobs)}
\newcommand{\MES}{\mathrm{MES}}
\newcommand{\CDPP}{\mathrm{CDPP}}

%elements
\newcommand{\Hel}{\mathsf{H}}
\newcommand{\Oel}{\mathsf{O}}
\newcommand{\Cel}{\mathsf{C}}
\newcommand{\nel}{b}
\newcommand{\bv}{\boldsymbol{b}}
\newcommand{\nv}{\boldsymbol{n}}
\newcommand{\fv}{\boldsymbol{f}}
\newcommand{\Jv}{\boldsymbol{J}}
\newcommand{\xv}{\boldsymbol{x}}
\newcommand{\yv}{\boldsymbol{y}}

%inference
\newcommand{\epsilonv}{\boldsymbol{\epsilon}}
\newcommand{\pv}{\boldsymbol{p}}
\newcommand{\qv}{\boldsymbol{q}}

%gp etc
\newcommand{\Dp}{D} % # of parameters
\newcommand{\Mp}{M} % # of models
\newcommand{\Kp}{K} % # of layers


\newcommand{\mv}{\boldsymbol{m}}
\newcommand{\muv}{\boldsymbol{\mu}}
\newcommand{\partialv}{\boldsymbol{\partial}}
\newcommand{\omegav}{\boldsymbol{\omega}}
\newcommand{\av}{\boldsymbol{a}}
\newcommand{\uv}{\boldsymbol{u}}
\newcommand{\vv}{\boldsymbol{v}}
\newcommand{\gv}{\boldsymbol{g}}
\newcommand{\tv}{\boldsymbol{t}}
\newcommand{\mva}{\boldsymbol{m}^\ast}
\newcommand{\Ng}{\mathcal{N}}
\newcommand{\zerov}{\boldsymbol{0}}


\newcommand{\Fv}{\boldsymbol{F}}
\newcommand{\zv}{\boldsymbol{z}}
\newcommand{\hv}{\boldsymbol{h}}
\newcommand{\evv}{\boldsymbol{e}}
\newcommand{\ntot}{n_\mathrm{tot}}
\newcommand{\lnntot}{q_\mathrm{tot}}
\newcommand{\lnnv}{\boldsymbol{q}}
\newcommand{\lambdav}{\boldsymbol{\lambda}}
\newcommand{\piv}{\boldsymbol{\pi}}
\newcommand{\Piv}{\boldsymbol{\Pi}}
\newcommand{\Lv}{\boldsymbol{L}}
\newcommand{\lambdatot}{\lambda_\mathrm{tot}}
\newcommand{\alphav}{\boldsymbol{\alpha}}
\newcommand{\betav}{\boldsymbol{\beta}}
\newcommand{\etav}{\boldsymbol{\eta}}
\newcommand{\nelements}{m}
\newcommand{\gtot}{g_\mathrm{tot}}


\newcommand{\Blam}{B_\nu}
\newcommand{\Ilam}{{\mathcal{I}_\nu}}
\newcommand{\Jlam}{{\mathcal{J}_\nu}}
\newcommand{\Jl}{{J_\nu}}
\newcommand{\Hl}{{H_\nu}}
\newcommand{\Kl}{{K_\nu}}
\newcommand{\JlVIS}{{J_{\nu}^\star}}
\newcommand{\HlVIS}{{H_{\nu}^\star}}
\newcommand{\KlVIS}{{K_{\nu}^\star}}
\newcommand{\Ilams}{{\mathcal{I}}}

\newcommand{\Fsum}{F_\mathrm{sum}}
\newcommand{\Fdif}{F_\mathrm{net}}
\newcommand{\dotFsum}{\dot{F}_\mathrm{sum}}
\newcommand{\dotFdif}{\dot{F}_\mathrm{net}}
\newcommand{\ddotFsum}{\ddot{F}_\mathrm{sum}}
\newcommand{\ddotFdif}{\ddot{F}_\mathrm{net}}

\newcommand{\Il}{I_\nu}
\newcommand{\Bl}{B_\nu}
\newcommand{\dBl}{\dot{B}_\nu}

\newcommand{\Jlu}{J_+}
\newcommand{\Hlu}{H_+}
\newcommand{\Jld}{J_-}
\newcommand{\Hld}{H_-}
\newcommand{\dHlu}{\dot{H}_+}
\newcommand{\dHld}{\dot{H}_-}
\newcommand{\Fnet}{F_\mathrm{net}}
\newcommand{\dFsum}{\dot{F}_\mathrm{sum}}
\newcommand{\ddFsum}{\ddot{F}_\mathrm{sum}}
\newcommand{\dFnet}{\dot{F}_\mathrm{net}}
\newcommand{\Bv}{\boldsymbol{\mathcal{B}}}
\newcommand{\Gv}{\boldsymbol{G}}
\newcommand{\Gaa}{\mathcal{G}_{11}}
\newcommand{\Gab}{\mathcal{G}_{12}}
\newcommand{\Gba}{\mathcal{G}_{21}}
\newcommand{\Gbb}{\mathcal{G}_{22}}
\newcommand{\Gcross}{\mathcal{G}_{\times}}



\setlength{\textwidth}{460pt}
\setlength{\voffset}{0cm}
\setlength{\topmargin}{0.0cm}
\setlength{\headheight}{0.0cm}
\setlength{\headsep}{1cm}
\setlength{\oddsidemargin}{0.0cm}
\setlength{\evensidemargin}{0.0cm}
\begin{document}
\title{Inferring Exoplanets}
\author{Hajime Kawahara}
\maketitle

\tableofcontents

\chapter{Introduction}
\input{en/introduction}

\input{en/quantities}

\chapter{Motion of A Planet}\label{ch:motion}
We consider quantities related to planetary motion as observable properties of exoplanets. Two such observables are the radial velocity curve, which measures the stellar motion along the line of sight induced by the planet, and astrometry, which traces the stellar motion projected on the celestial sphere.

\input{en/dynamics}
\input{en/radial_velocity}

\chapter{Inference of Motion}\label{ch:infer}

In Chapter~\ref{ch:motion}, we discussed physical models of observables related to the motion of exoplanets. All of these are one-dimensional time-series data. In this chapter, we focus on estimating the internal parameters—namely the physical properties of exoplanets—by combining physical and statistical models with one-dimensional time-series data.

\input{en/inference}
\input{en/bayes}
\input{en/rvbayes}
\input{en/gp}

\chapter{Spectra from Exoplanets}

In this chapter, we provide an overview of the signals that encode information intrinsic to the planet itself. Planetary signals can be broadly categorized into three types: starlight transmitted through the planetary atmosphere, thermal emission from the planet, and starlight reflected by the planet, as listed in  Table~\ref{tab:spec_sep}. By performing spectroscopy of these planetary signals, we can investigate the atmospheres of exoplanets. 
\begin{table*}[htbp]
\footnotesize
\centering
\caption{Classification of Spectral Types}
\label{tab:spec_sep}
\begin{tabular}{ccc}
\hline
Spectrum & Technique & Main Current Targets \\
\hline
Transmission & Transmission Spectroscopy & Transiting Systems \\
             & High-Dispersion Spectroscopy & Transiting Systems \\
\hline
Emission & Secondary Eclipse Spectroscopy & Transiting Systems \\
         & Direct Imaging & Young Planets \\
         & High-Dispersion Spectroscopy & Bright Systems \\
\hline
Reflection & Direct Imaging & Nearby Earth-like Planets \\
           & High-Dispersion Spectroscopy & Bright Systems \\
\hline
\end{tabular}
\end{table*}
In all cases, the light passing through the planetary atmosphere is modified by absorption and scattering due to atmospheric components such as molecules, atoms, and clouds. Detecting these changes in the spectrum allows us to infer the properties of the atmosphere. Depending on the pathway of the light, the signals can be classified into those that use radiation emitted by the planet itself and those that rely on starlight. The former will hereafter be referred to simply as \textit{emission}, while the latter correspond to transit transmission and reflection. These latter two are often regarded as \textit{passive} methods, as the planetary signal is imprinted on the starlight. 

\begin{table*}[htbp]
\footnotesize
\centering
\caption{Classification of Exoplanet Spectroscopic Methods}
\label{tab:light_sep}
\begin{tabular}{ccc}
\hline
Separation Method & Technique & Applicable Targets \\
\hline
Data Analysis & Transmission Spectroscopy & Transiting Planets \\
              & Secondary Eclipse Spectroscopy & Transiting Planets \\
              & High-Dispersion Spectroscopy & Transiting or Hot Planets \\
\hline
Instrumentation & Direct Imaging & Young Planets or Nearby Earth-like Planets \\
\hline
\end{tabular}
\end{table*}



The primary challenge in exoplanet spectroscopy lies in separating and removing the overwhelming starlight in order to extract the planetary spectrum. Table~\ref{tab:light_sep} shows the classification of techniques based on their separation method.  

In the first category, separation is achieved through data analysis. Here, spectra are obtained in which starlight and planetary light are not physically separated, and the planetary signal is extracted through data processing. In transiting systems, the reduction in brightness during transit allows us to measure the planetary radius. By carrying this out as a function of wavelength, one can obtain atmospheric information. This technique is called \underline{transmission spectroscopy}. In transiting systems, when the planet passes behind the star, the light from the planet is blocked, producing another dip in brightness. This is known as the \underline{secondary eclipse}, and it can be used to isolate the planetary emission. These two techniques are applicable only to transiting systems.  

Next, for spectra in which starlight and planetary light are blended, \underline{high-dispersion spectroscopy} can be used to extract the planetary signal. This method exploits the difference in radial velocity between the star and the planet to identify spectral features originating from the planet, in particular the absorption lines of molecules and atoms in its atmosphere. Importantly, this technique can be applied to non-transiting systems as well.  

Finally, there is the instrumental method of suppressing starlight through \underline{direct imaging}. In this case, detectability is determined by the contrast ratio between the star and the planet, as well as their angular separation. The closer the system is to the observer, the larger the apparent separation, making nearby planetary systems favorable targets. Current direct imaging capabilities are limited to detecting systems with moderate contrasts, and thus detections have been restricted to young, self-luminous planets. However, with future advances in space-based direct imaging and improvements in ground-based instrumentation, the detection of more typical planetary systems will in principle become possible.  

\input{en/transit}
\input{en/emisreflect}
\input{en/spectrum_example}

\chapter{Exoplanet Atmosphere}
\input{en/chemistry}
\input{en/molecule}
\input{en/atmosphere}

\chapter{Atmospheric Inference from Spectra}

The overall workflow for inferring atmospheric properties from spectra is as follows.
\begin{enumerate}
    \item Set up an atmospheric forward model: construct a parametric atmospheric model and implement code that generates a spectrum from the model.
    \item Specify a probabilistic model: define the error model associated with the spectral data and the prior distributions for the atmospheric model.
    \item Perform Bayesian inference: compute the likelihood from the observed spectrum and the model-generated spectrum, and obtain the posterior distributions of the parameters in the atmospheric model.
\end{enumerate}

In the most agnostic ``free-retrieval'' approach - i.e., one that includes minimal physico-chemical assumptions—the atmospheric forward model treats the temperature–pressure (T–P) profile itself as an unknown to be retrieved. One may assume an appropriate parametrized profile, or adopt a more flexible profile using a Gaussian process. Molecular abundances can likewise be treated as parameters. For high-temperature atmospheres, one can instead assume thermochemical equilibrium and retrieve elemental abundances.

Once the atmospheric forward model has been specified, the next step is spectrum generation. Below, we describe this spectrum-synthesis component in detail. The Bayesian inference procedure is essentially the same as in Chapter~\ref{ch:infer}.


\input{en/transmission}
\input{en/emisspectrum}
\input{en/twostream}




%\input{mcmc}
%\input{abc}

\footnotesize
\bibliography{ref}{}

\subsection*{Textbooks}

For Bayesian estimation, the standard reference is Gelman et al.'s \textit{Bayesian Data Analysis} (BDA; Morikita Publishing). Kevin P.~Murphy's \textit{Probabilistic Machine Learning} offers an even more comprehensive treatment of key topics.


\bibliographystyle{plain}

\end{document}

